From physical principles to ???

### 2016-01-14 2015CompleteDerivation/completeDerivation.tex
More on state variables

Third, operationally we must reliably prepare and measure states. That is, we need a process for which the input settings of our preparing device determine the outgoing state of the system; and a process for which the incoming state of the system can be reconstructed by the output of the measuring device. That is, our system must participate in a deterministic and reversible process with the preparing and measuring device. How else could we calibrate our experimental apparatus?

This link between state definition and deterministic processes should not be too surprising as the state, in the context of thermodynamics and system theory, is often defined as \emph{the set of variables needed to determine the future evolution of the system}. As we saw before, this applies also to statistical processes: the distribution (the ensemble) as a whole can be indeed calculated, measured and prepared. We can also describe the evolution of each element of the distribution provided that: we have a way to isolate it and study it under deterministic and reversible motion (so that we can define its states); the non-deterministic motion does not alter the system (the set of states remains the same). For example, consider a wooden ball described by position and momentum of the center of mass: if randomly scattered by other balls retain the set of states, if thrown into a fire it does not.

@@

To identify and name states, we will typically use a set of quantities, typically numbers. For example, the orbital of an electron in a hydrogen item is identified by the quantum numbers $n$, $l$, $m$ and $s$. We call each of these quantities \emph{state variables} and \emph{state vector} a set of state variables that can identify one and only one state. We avoid the term coordinates, even though that's what they are mathematically, as it will create confusion with spatial coordinates. We also avoid the term observable or measurable, as not all state variables map directly to the outcome of an experiment (e.g. conjugate momentum as it is not gauge invariant). It's clear, though, that even in those cases there must exist, at least in principle, a way to distinguish between the possible values of the state variable (e.g once we fix the vector potential, conjugate momentum can be measured via the kinetic momentum. Therefore we can assume that all state variables are in the end physically distinguishable, yet we do not assume all state variables have equal standing in physical meaning. Their purpose is just to label states, and the chosen name is a good reminder.

@@

When assembling multiple state variable into a state vector, it is important to understand how they relate to each other. Consider the orbital of an electron in a hydrogen item, which is identified by the quantum numbers $n$, $l$, $m$ and $s$. For each combination of $n$, $l$ and $m$, the spin $s$ can have to value. The choice of spin is independent from the rest. The choice of $l$, though, depends on the choice of $n$: for $n=1$ only $l=s$ is available; for $n=2$ we can choose $l=s$ or $l=p$. The choices are not independent. That is: two or more variables are independent if there always exist a state for any possible combination.

>>>>

Third, operationally we must reliably prepare and measure states. That is, we need a process for which the input settings of our preparing device determine the outgoing state of the system; and a process for which the incoming state of the system can be reconstructed by the output of the measuring device. That is, our system must participate in a deterministic and reversible process with the preparing and measuring device. Without it we wouldn't be able to calibrate our experimental apparatus.

This link between state definition and deterministic processes should not be too surprising as the state, in the context of thermodynamics and systems theory, is often defined as \emph{the set of variables needed to determine the future evolution of the system}. As we saw before, this applies also to statistical processes: the distribution (the ensemble) as a whole can be indeed calculated, measured and prepared. We can also describe the evolution of each element of the distribution provided that: we have a way to isolate it and study it under deterministic and reversible motion (so that we can define its states); the non-deterministic motion does not alter the system (the set of states remains the same). For example, consider a wooden ball described by position and momentum of the center of mass: if randomly scattered by other balls retain the set of states, if thrown into a fire it does not.

@@

To identify and name states, we will typically use a set of quantities, typically numbers. For example, the orbital of an electron in a hydrogen item is identified by the quantum numbers $n$, $l$, $m$ and $s$. We call each of these quantities \emph{state variable} and \emph{state vector} a set of state variables that can identify one and only one state. We avoid the term coordinates, even though that's what they mathematically are, as it will create confusion with space coordinates. We also avoid the term observable or measurable, as not all state variables map directly to the outcome of an experiment (e.g. conjugate momentum as it is not gauge invariant). It's clear, though, that even in those cases there must exist, at least in principle, a way to distinguish between the possible values of the state variable (e.g once we fix the vector potential, conjugate momentum can be measured via the kinetic momentum). Therefore we can assume that all state variables are in the end physically distinguishable, yet we do not assume all state variables have equal standing in physical meaning. Conversely, some measurable quantities are not directly state variables: they may not be defined, as states are, at constant time (e.g. velocity) or may depend on the process at hand (e.g. acceleration). Yet, once a process is fully defined (e.g. a Hamiltonian is given), we may have a relationship between those observables and state variables. The purpose of state variables is just to label states, which are physically distinguishable and defined at a particular time. Therefore the state variable will be physically distinguishable and defined at constant time. But there is nothing telling us that they must be directly physically meaningful or that their relationship to measurable quantities is the same if we change physical processes.

@@

To keep track of the state evolution, we'll need to understand how state variables change under continuous and reversible evolution. The first approach is to study the change in the state variables values from $q^i(\mathcal{s})$ to $q^i(\hat{\mathcal{s}})$. This is especially useful when state variables correspond to physically meaningful quantities. For example, we track how the temperature or the pressure of an ideal gas has changed. However, what we'll want to do is to find relationships that are common for all deterministic and reversible processes, which is hard to do in this picture. The second approach is to evolve the state variables as well. The idea is to keep the connection to the initial state by labelling the future state, instead by the future values of $q^i$, by the original one. So we introduce a new set of variables for which $\hat{q}^i(\hat{\mathcal{s}})=q^i(\mathcal{s})$. We can always do this because we have bijective correspondence between initial and final states. This makes it easier to study relationships that are invariant during the evolution: once we verify them for $q^i$ on the initial states, they will automatically hold for the $\hat{q}^i$ at later (or previous) times. Naturally, as $\hat{q}^i$ are new variables that will not hold the same physical meaning. For example, each will be a combination of pressure and temperature that depends on the particular process. We'll typically use the first picture to describe equations of motions and the second to describe invariants. Note, though, that the two picture are tightly linked. In fact:
\begin{align*}
\Delta q^i &= q^i(\hat{\mathcal{s}}) - q^i(\mathcal{s}) \\
\hat{q}^i(\hat{\mathcal{s}}) &= q^i(\hat{\mathcal{s}}) - \Delta q^i
\end{align*}

\begin{prop}\label{evolved variable}
TODO: Define and prove existence
\end{prop}

When assembling multiple state variable into a state vector, it is important to understand how they relate to each other. Consider the orbital of an electron in a hydrogen item, which is identified by the quantum numbers $n$, $l$, $m$ and $s$. For each combination of $n$, $l$ and $m$, the spin $s$ can have to value. The choice of spin is independent from the rest. The choice of $l$, though, depends on the choice of $n$: for $n=1$ only $l=s$ is available; for $n=2$ we can choose $l=s$ or $l=p$. The choices are not independent. That is: two or more variables are independent if there always exist a state for any possible combination, if the total number of states is the cartesian product of the possibilities of each variable. For discrete space variables,
